# RNA-seq data analysis with DESeq2

Authors:
    Michael I. Love^[UNC-Chapel Hill, NC, US],
    Simon Anders^[ZMBH Heidelberg, Germany],
    Wolfgang Huber^[EMBL Heidelberg, Germany]
Last modified: 25 June, 2018.

<!-- 
  bookdown::render_book("201_Love_DESeq2.Rmd", "bookdown::gitbook") 
-->

## Overview

### Description

In this workshop, we will give a quick overview of the most useful
functions in the DESeq2 package, and a basic RNA-seq analysis. We will
cover: how to quantify transcript expression from FASTQ files using
Salmon, import quantification from Salmon with tximport and tximeta,
generate plots for quality control and exploratory data analysis EDA
(also using MultiQC), perform differential expression (DE) (also using
apeglm), overlap with other experimental data (using AnnotationHub),
and build reports (using ReportingTools and Glimma). We will give a
short example of integration of DESeq2 with the zinbwave package for
single-cell RNA-seq differential expression. The workshop is designed
to be a lab with plenty of time for questions throughout the lab. 

### Pre-requisites

* Basic knowledge of R syntax

Non-essential background reading:

* DESeq2 paper: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4302049/>
* tximport paper: <https://f1000research.com/articles/4-1521/v2>
* apeglm paper: <https://www.biorxiv.org/content/early/2018/04/17/303255>

### Participation

Students will participate by following along an Rmarkdown document,
and asking questions throughout the workshop.

### _R_ / _Bioconductor_ packages used

* DESeq2
* tximport
* apeglm
* AnnotationHub
* ReportingTools
* Glimma
* zinbwave

### Time outline

| Activity                      | Time |
|:------------------------------|:-----|
| Overview of packages          | 20m  |
| Quantification and import     | 20m  |
| EDA and DE                    | 20m  |
| Downstream analysis & reports | 20m  |
| ZINB-WaVE integration         | 20m  |
| Additional questions          | 20m  |

### Workshop goals and objectives

Learning goals

* Visually assess quality of RNA-seq data 
* Perform basic differential analysis of RNA-seq data 
* Compare RNA-seq results with other experimental data

Learning objectives

* Quantify transcript expression from FASTQ files
* Import quantification into R/Bioconductor
* Perform quality control and exploratory data analysis
* Perform differential expression
* Overlap with other experimental data
* Build dynamic reports
* Integrate DESeq2 and zinbwave for single-cell RNA-seq data

## Preparing data for *DESeq2*

### Experimental data

The data used in this workflow is stored in the *airway* package that
summarizes an RNA-seq experiment wherein airway smooth muscle cells
were treated with dexamethasone, a synthetic glucocorticoid steroid
with anti-inflammatory effects [@Himes2014RNASeq]. Glucocorticoids
are used, for example, by people with asthma to reduce inflammation of
the airways. In the experiment, four primary human airway smooth
muscle cell lines were treated with 1 micromolar dexamethasone for 18
hours. For each of the four cell lines, we have a treated and an
untreated sample. For more description of the experiment see the
[PubMed entry 24926665](http://www.ncbi.nlm.nih.gov/pubmed/24926665)
and for raw data see the
[GEO entry GSE52778](http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE52778).

We will show how to import RNA-seq quantification data using an
alternative dataset (the *tximportData* package which is used in the
*tximport* vignette). Afterward we will load counts for the *airway*
dataset, which were counted using *summarizeOverlaps* from the
*GenomicAlignments* package. As described below, we recommend the
*tximport* pipeline for producing count matrices, but we do not yet
have a Bioconductor package containing the necessary quantification
files for the *airway* dataset.

### Modeling count data

As input, the count-based statistical methods, such as *DESeq2*
[@Love2014Moderated], *edgeR* [@Robinson2009EdgeR], *limma* with the
voom method [@Law2014Voom], *DSS* [@Wu2013New], *EBSeq*
[@Leng2013EBSeq] and *baySeq* [@Hardcastle2010BaySeq], expect input
data as obtained, e.g., from RNA-seq or another high-throughput
sequencing experiment, in the form of a matrix of counts.  The value
in the *i*-th row and the *j*-th column of the matrix tells how many
reads (or fragments, for paired-end RNA-seq) have been assigned to
gene *i* in sample *j*. Analogously, for other types of assays, the
rows of the matrix might correspond e.g., to binding regions (with
ChIP-Seq), species of bacteria (with metagenomic datasets), or peptide
sequences (with quantitative mass spectrometry).

The values in the matrix should be counts of sequencing
reads/fragments. This is important for the statistical models used by
*DESeq2* and *edgeR* to hold, as only counts allow assessing the
measurement precision correctly. It is important to not provide counts
that were pre-normalized for sequencing depth (also called library
size), as the statistical model is most powerful when applied to
un-normalized counts and is designed to account for library size
differences internally.

### Transcript abundances

In this workflow, we will show how to use transcript abundances as
quantified by the [Salmon](https://combine-lab.github.io/salmon/)
[@Patro2017Salmon] software package. *Salmon* and other methods, such
as [Sailfish](http://www.cs.cmu.edu/~ckingsf/software/sailfish/)
[@Patro2014Sailfish],
[kallisto](https://pachterlab.github.io/kallisto/) [@Bray2016Near],
or [RSEM](http://deweylab.github.io/RSEM/) [@Li2011RSEM], estimate the
relative abundances of all (known, annotated) transcripts without
aligning reads. Because estimating the abundance of the transcripts
involves an inference step, the counts are *estimated*. Most methods
either use a statistical framework called Estimation-Maximization or
Bayesian techniques to estimate the abundances and counts.  Following
quantification, we will use the *tximport* [@Soneson2015Differential]
package for assembling estimated count and offset matrices for use
with Bioconductor differential gene expression packages.

The advantages of using the transcript abundance quantifiers in
conjunction with *tximport* to produce gene-level count matrices and
normalizing offsets, are: 

1. this approach corrects for any potential changes in gene length
across samples (e.g. from differential isoform usage)
[@Trapnell2013Differential] 
2. some of these methods are substantially faster and require less
memory and less disk usage compared to alignment-based methods 
3. it is possible to avoid discarding those fragments that can align
to multiple genes with homologous sequence [@Robert2015Errors]. 

Note that transcript abundance quantifiers skip the generation of
large files which store read alignments (SAM or BAM files), instead
producing smaller files which store estimated abundances, counts and
effective lengths per transcript. For more details, see the manuscript
describing this approach [@Soneson2015Differential] and the *tximport*
package vignette for software details.

A full tutorial on how to use the *Salmon* software for quantifying
transcript abundance can be
found [here](https://combine-lab.github.io/salmon/getting_started/).

### *Salmon* quantification

We begin by providing *Salmon* with the sequence of all of the
reference transcripts, which we will call the *reference
transcriptome*. We recommend to use the GENCODE human
transcripts, which can be downloaded from the
[GENCODE website](https://www.gencodegenes.org/). 
On the command line, creating the transcriptome index looks like:

```
salmon index -i gencode.v99_salmon_0.10.0 -t gencode.v99.transcripts.fa.gz
```

The `0.10.0` refers to the version of *Salmon* that was used, and is
useful to put into the index name.

To quantify an individual sample, `sample_01`, the following command
can be used:

```
salmon quant -i gencode.v99_salmon_0.10.0 -p 6 --libType A \
  --gcBias --biasSpeedSamp 5 \
  -1 sample_01_1.fastq.gz -2 sample_01_2.fastq.gz \
  -o sample_01
```

In simple English, this command says to "quantify a sample using this
transcriptome index, with 6 threads, using automatic 
[library type](http://salmon.readthedocs.io/en/latest/library_type.html) detection,
using GC bias correction (the bias speed part is now longer
needed with current versions of *Salmon*), here are the first and second
read, and use this output directory." The output directory will be
created if it doesn't exist, though if earlier parts of the path do
not exist, it will give an error. A single sample of human RNA-seq
usually takes ~5 minutes with the GC bias correction.

Rather than writing the above command on the command line multiple
times for each sample, it is possible to loop over files using a
bash loop, or more advanced workflow management
systems such as Snakemake [@Koster2012Snakemake] or Nextflow
[@Di2017Nextflow].

## Importing into R with *tximport*

Following quantification, we can use *tximport* to import the data
into R and perform statistical analysis using Bioconductor packages.
Normally, we would simply point *tximport* to the `quant.sf` files on
our machine. However, because we are distributing these files as part
of an R package, we have to do some extra steps, to figure out where
the R package, and so the files, are located on *your* machine.

We will show how to import *Salmon* quantification files using the
data in the *tximportData* package. The quantified samples are six
samples from the [GEUVADIS Project](http://www.geuvadis.org/web/geuvadis) [@Lappalainen].
The output directories from the above *Salmon* quantification calls has been
stored in the `extdata` directory of the *tximportData* package.
The R function *system.file* can be used to find out where on your
computer the files from a package have been installed. Here we ask for
the full path to the `extdata` directory, where R packages store
external data, that is part of the *tximportData* package.

```{r}
library("tximportData")
dir <- system.file("extdata", package="tximportData")
list.files(dir)
```

The *Salmon* quantification directories are in the `salmon` directory. 

```{r}
list.files(file.path(dir,"salmon"))
```

The identifiers used here are the *ERR* identifiers from the 
[European Nucleotide Archive](https://www.ebi.ac.uk/ena). 
We need to create a named vector pointing to the quantification
files. We will create a vector of filenames first by reading in a
table that contains the sample IDs, and then combining this with `dir`
and `"quant.sf.gz"`. (We gzipped the quantification files to make the
data package smaller, this is not a problem for R functions that we
use to import the files.)

```{r}
samples <- read.table(file.path(dir,"samples.txt"), header=TRUE)
samples
files <- file.path(dir, "salmon", samples$run, "quant.sf.gz")
names(files) <- paste0("sample",1:6)
all(file.exists(files))
```

Transcripts need to be associated with gene IDs for gene-level
summarization. We therefore will construct a *data.frame* called
`tx2gene` with two columns: 1) transcript ID and 2) gene ID. The
column names do not matter but this column order must be used. The
transcript ID must be the same one used in the abundance files. This
can most easily be accomplished by downloading the GTF file at the
same time that the transcriptome FASTA is downloaded, and generating
`tx2gene` from the GTF file using Bioconductor's *TxDb*
infrastructure.

Generating a *TxDb* from a GTF file can be easily accomplished with
the *makeTxDbFromGFF* function. This step requires a few minutes of
waiting, and a large file. We therefore skip this step, but show the
code that is used to create the `tx2gene` table, assuming the correct
*TxDb* object has been created.

Creating the `tx2gene` *data.frame* can be accomplished by calling the
*select* function from the *AnnotationDbi* package on a *TxDb* object.
The following code could be used to construct such a table:

```{r}
library("TxDb.Hsapiens.UCSC.hg38.knownGene")
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
k <- keys(txdb, keytype="TXNAME")
tx2gene <- select(txdb, k, "GENEID", "TXNAME")
```

In this case, we've used the Gencode v27 CHR transcripts to build our
*Salmon* index, and we used `makeTxDbFromGFF` and code similar to the chunk
above to build the `tx2gene` table. We then read in a pre-constructed
`tx2gene` table:

```{r}
library(readr)
tx2gene <- read_csv(file.path(dir, "tx2gene.gencode.v27.csv"))
head(tx2gene)
```

Finally the following line of code imports *Salmon* transcript
quantifications into R, collapsing to the gene level using the
information in `tx2gene`.

```{r}
library("tximport")
library("jsonlite")
library("readr")
txi <- tximport(files, type="salmon", tx2gene=tx2gene)
```

The `txi` object is simply a list of matrices (and one character
vector):

```{r}
names(txi)
txi$counts[1:3,1:3]
txi$length[1:3,1:3]
txi$abundance[1:3,1:3]
txi$countsFromAbundance
```

If we were continuing with the GEUVADIS samples, we would then create
a *DESeqDataSet* with the following line of code. Because there are no
differences among the samples (same population and same sequencing
batch), we specify a *design formula* of ~1, meaning we can only fit
an intercept term -- so we cannot perform differential expression
analysis with these samples.

```{r}
library("DESeq2")
dds <- DESeqDataSetFromTximport(txi, samples, ~1)
dds$center
dds$pop
```

## Exploratory data analysis

We will now switch over to the *airway* experiment, counts of which
are already prepared in a *SummarizedExperiment* object. In this case,
the object that we load is the output of the *summarizeOverlaps*
function in the *GenomicAlignments* package, and the exact code used
to produce this object can be seen by typing `vignette("airway")` into
the R session, to pull up the *airway* software vignette. There are
multiple ways to produce a count table and import it into *DESeq2*,
and these are summarized in this section of the
[RNA-seq gene-level workflow](https://bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html#deseq2-import-functions).

```{r}
library("airway")
data("airway")
```

We want to specify that `untrt` is the reference level for the dex
variable: 

```{r}
airway$dex <- relevel(airway$dex, "untrt")
airway$dex
```

We can quickly check the millions of fragments that uniquely aligned
to the genes (the second argument of *round* tells how many decimal
points to keep).

```{r}
round( colSums(assay(airway)) / 1e6, 1 )
```

We can inspect the information about the samples, by pulling out the
`colData` slot of the *SummarizedExperiment*:

```{r}
colData(airway)
table(airway$cell)
table(airway$dex)
```

If we had not already loaded *DESeq2*, we would do this, and then
create a *DESeqDataSet*. We want to control for the cell line, while
testing for differences across dexamethasone treatment, so we use a
design of `~ cell + dex`:

```{r}
library("DESeq2")
dds <- DESeqDataSet(airway, design = ~ cell + dex)
```

We will perform a minimal filtering to reduce the size of the
dataset. We do not need to retain genes if they do not have a count of
5 or more for 4 or more samples as these genes will have no
statistical power to detect differences, and no information to compute
distances between samples.

```{r}
keep <- rowSums(counts(dds) >= 5) >= 4
table(keep)
dds <- dds[keep,]
```

Some very basic exploratory analysis is to examine a boxplot of the
counts for each sample. We will take the logarithm so that large
counts do not dominate the boxplot:

```{r}
boxplot(log10(counts(dds)+1))
```

The main function in *DESeq2* involves computation of *size factors*
which normalize for differences in sequencing depth among samples. We
can also compute these size factors manually, so that the *normalized
counts* are available for plotting:

```{r}
dds <- estimateSizeFactors(dds)
boxplot(log10(counts(dds,normalized=TRUE)+1))
```

Taking the logarithm of counts plus a pseudocount of 1 is a common
transformation, but it tends to inflate the sampling variance of low
counts such that it is even larger than biological variation across
groups of samples. In *DESeq2* we therefore provide transformations
which produce log-scale data such that the systematic trends have been
removed. Our recommended transformation is the variance-stabilizing
transformation, or VST, and it can be called with the *vst* function:

```{r}
vsd <- vst(dds)
class(vsd)
```

This function does not return a *DESeqDataSet*, because it does not
return counts, but instead continuous values (on the log2 scale).
We can access the transformed data with *assay*:

```{r}
assay(vsd)[1:3,1:3]
```

The VST data is appropriate for calculating distances between samples
or for performing PCA. More information about PCA and distance
calculation can be found in the
[RNA-seq gene-level workflow](https://bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html#deseq2-import-functions).
In short, PCA plots allow us to visualize the most dominant axes of
variation in our data, which is useful for both quality control, and
to get a sense of how large the inter-sample differences are across
and within conditions. Here we see that PC1 (the primary axis of
variation in the data) separates the treated and untreated samples:

```{r}
plotPCA(vsd, "dex")
```

With some additional *ggplot2* code, we can also indicate which
samples belong to which cell line:

```{r}
library("ggplot2")
pcaData <- plotPCA(vsd, intgroup = c( "dex", "cell"), returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(x = PC1, y = PC2, color = dex, shape = cell)) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed()
```

Note that we do not recommend working with the transformed data for
the primary differential expression analysis. Instead we will use the
original counts and a *generalized linear model* (GLM) which takes
into account the expected variance from either low or high counts.
For statistical details, please refer to the *DESeq2* methods paper
[@Love2014Moderated].

## Differential expression analysis

### Standard DE steps

Differential expression analysis in *DESeq2* is performed by calling
the following two functions:

```{r}
dds <- DESeq(dds)
res <- results(dds)
```

The results table `res` contains the results for each gene (in the
same order as in the *DESeqDataSet*). If we want to see the top genes,
we can order it like so:

```{r}
head(res[order(res$pvalue),])
```

We can plot the counts for the top gene using `plotCounts`:

```{r}
plotCounts(dds, which.min(res$pvalue), "dex")
```

We can examine all the log2 fold changes (LFC) due to dexamethasone
treatment over the mean of counts using `plotMA`:

```{r}
plotMA(res, ylim=c(-5,5))
```

Note that there are many large LFC which are not significant (grey
points) on the left side of the MA-plot above. These obtain a large
LFC because of the imprecision of log counts. For more informative
visualization and more accurate ranking of genes by effect size (the
log fold change may sometimes be referred to as an *effect size*), we
recommend to use *DESeq2*'s functionality for shrinking LFCs. Our most
recent methodological development is the *apeglm* shrinkage estimator,
which is available in *DESeq2*'s *lfcShrink* function:

```{r}
library("apeglm")
resultsNames(dds)
res2 <- lfcShrink(dds, coef="dex_trt_vs_untrt", type="apeglm")
```

```{r}
par(mfrow=c(1,2))
plotMA(res, ylim=c(-3,3), main="No shrinkage")
plotMA(res2, ylim=c(-3,3), main="apeglm")
```

### Minimum effect size

If we don't want to report as significant genes with small LFC, we can
specify a minimum *biologically meaningful* effect size, by choosing
an LFC and testing against this. We can either perform such a
threshold test using the unshrunken LFCs or the LFCs provided by
*lfcShrink* using the *apeglm* method:

```{r}
res.lfc <- results(dds, lfcThreshold=1)
res.lfc2 <- lfcShrink(dds, coef="dex_trt_vs_untrt", type="apeglm",
                      lfcThreshold=1)
```

Note that *testing* against an LFC threshold is not equivalent to
testing against a null hypothesis of 0 and then filtering on LFC
values. We prefer the former, as discussed in @Love2014Moderated and
@Zhu2018. 

The *apeglm* method provides s-values [@Stephens2016] when
`svalue=TRUE` or when we supply a minimum effect size as above. These
are analogous to q-values or adjusted p-values, in that the genes with
s-values less than $\alpha$ should have an aggregate rate of false
sign or being smaller in absolute value than our given LFC threshold,
which is bounded by $\alpha$.

```{r}
par(mfrow=c(1,2))
plotMA(res.lfc, ylim=c(-5,5), main="No shrinkage, LFC test")
plotMA(res.lfc2, ylim=c(-5,5), main="apeglm, LFC test", alpha=0.01)
```

## *AnnotationHub*

We will use the *AnnotationHub* package to attach additional
information to the results table. *AnnotationHub* provides an
easy-to-use interface to more than 40,000 annotation records. A
record may be peaks from a ChIP-seq experiment from ENCODE, the
sequence of the human genome, a *TxDb* containing information about
transcripts and genes, or an *OrgDb* containing general information
about biological identifiers for a particular organism.

```{r}
library("AnnotationHub")
ah <- AnnotationHub()
```

The following code chunk, un-evaluated here, launches a browser for
navigating all the records available through *AnnotationHub*.

```{r eval=FALSE}
display(ah)
```

We can also query using keywords with the *query* function:

```{r}
query(ah, c("OrgDb","Homo sapiens"))
```

To pull down a particular record we use double brackets and the *name*
of the record:

```{r}
hs <- ah[["AH61777"]]
hs
```

The *rownames* of the results table are Ensembl IDs, and most of these
are entries in *OrgDb* (although thousands are not).

```{r}
columns(hs)
table(rownames(res) %in% keys(hs, "ENSEMBL"))
```

We can use the *mapIds* function to add gene symbols, using `ENSEMBL`
as the keytype, and requesting the column `SYMBOL`. 

```{r}
res$symbol <- mapIds(hs, rownames(res), column="SYMBOL", keytype="ENSEMBL")
head(res)
```

## Building reports

```{r}
library(ReportingTools)
tmp <- tempdir()
rep <- HTMLReport(shortName="airway", title="Airway DGE", basePath=tmp, reportDirectory="report")
publish(res, rep, dds, n=20, make.plots=TRUE, factor=dds$dex)
finish(rep)
browseURL(file.path(tmp,"report","airway.html"))
```

TODO: glimma

## Integration with *ZINB-WaVE*

TODO
