# Working with Genomic Data in R with the DECIPHER package

Authors:
    Nicholas Cooley^[University of Pittsburgh]
    Erik Wright^[University of Pittsburgh]
Last Modified: 29 June, 2018
    
<!--
  bookdown::render_book("", "bookdown::gitbook")
-->

## Overview

### Workshop Description

In this workshop we will give an introduction to working with biological
sequence data in R using the Biostrings and DECIPHER packages. We will cover:

* Importing, viewing, and manipulating genomes in R
* Construction of sequence databases for organizing genomes
* Mapping syntenic regions between genomes with the FindSynteny function
* Understanding, accessing, and viewing objects of class Synteny
* Using syntenic information to predict orthologous genes
* Alignment of sequences and genomes with DECIPHER
* Downstream analyses enabled by syntenic mapping

### Pre-requisites

* Familiarity with Biostrings
* Familiarity with DECIPHER Databases (Ref. 1)

1. [Wright, E. S. The R Journal 2016, 8 (1), 352–359.](https://journal.r-project.org/archive/2016-1/wright.pdf)

### Workshop Participation

This will be a lab where participants follow along on their computers.

### _R_ / _Bioconductor_ packages used

* Biostrings
* DECIPHER

### Time outline

| Activity                          | Time |
|-----------------------------------|------|
| Packages/Introduction             | 5m   |
| Basic commands for sequences      | 5m   |
| Sequence databases                | 10m  |
| Demonstration of synteny mapping  | 10m  |
| Explanation of function arguments | 10m  |
| Dissecting Synteny objects        | 10m  |
| Visualization of syntenic blocks  | 10m  |
| Alignment of syntenic regions     | 10m  |
| Ortholog prediction from Synteny  | 10m  |
| Constructing phylogenetic trees   | 10m  |

## Workshop goals and objectives

### Learning goals

* Understand a simple workflow for analysis of sequences in R and DECIPHER
* Learn the basic use and appropriate application of functions within DECIPHER

### Learning objectives

* Learn basic commands for working with sequences in R
* Import genomes from online repositories or local files
* Map synteny between genomes
* Analyze a synteny map among multiple genomes
* Develop an understanding of the data structures involved
* Predict orthologs from syntenic maps
* Select core and pan genomes from predicted orthologs
* Construct and interpret phylogenetic trees

### Workshop

This is a workflow...
```{r, echo = FALSE}
# Load packages
suppressMessages(library(DECIPHER))
suppressMessages(library(phytools))
suppressMessages(library(FindHomology))
load("/Users/npc19/Desktop/WorkshopData.RData",
     verbose = TRUE)
```
## selection of genomes, deposition into sql database
```{r, echo = FALSE}
# Decipher works in sql databases, and our genomes can be sent directly to this database from the ftp site
DBPath <- tempfile() # in your own work a temp file is likely inadvisable

# Genomes is a file of character vectors of ftp addresses
for (i in seq_along(Genomes)) {
  Seqs2DB(seqs = Genomes[i],
          type = "FASTA",
          dbFile = DBPath,
          identifier = as.character(i),
          tblName = "Seqs")
}
```
## Comparison of genomes
```{r, echo = FALSE}
# In the interest of time we will use the first 4
SyntenyObject <- FindSynteny(dbFile = DBPath,
                             identifier = as.character(1:4),
                             verbose = FALSE)
# Plot SyntenyObject one way
plot(SyntenyObject,
     type = "neighbor")
# Plot a Subset of SyntenyObject another way
pairs(SyntenyObject[2:3, 2:3])
# Plot all of SyntenyObject in a pairs plot
pairs(SyntenyObject)

# Utilizing gene calls, and the hits from the synteny object
MatrixObject <- NucleotideOverlap(SyntenyObject = SyntenyObject,
                                  GeneCalls = GeneCalls[1:4],
                                  Verbose = FALSE)

# This object provides a list of genes that are paired by syntenic hits for each pair of genomes
MatrixObject[1, 2][[1]][1:10, ]
```
## Predict Homology between genes that are linked by syntenic hits
```{r, echo = FALSE}
Homologs <- Catalog(MatrixObject,
                    Verbose = FALSE)

# This object is a list of matrices that describes which genes -represented by integers- are paired, across which pairs of genomes, represented by columns
length(Homologs)
Homologs[[1]]
Homologs[head(which(sapply(Homologs,
                           function(x) nrow(x)) == 1L))]

# In this case, this is trivial, but in cases with larger sets of genomes, it may not be
MaxRows <- max(sapply(Homologs,
                      function(x) nrow(x)),
               na.rm = TRUE)
# Genes that are homologous across the whole set of genomes, and are completely transitive, represent the 'core' genome for this set of genomes
CoreSet <- which(sapply(Homologs,
                        function(x) nrow(x)) == MaxRows)
```
Use the core genome to create an alignment
```{r, echo = FALSE}
# Select core genes individually from each genome and align them, then concatonate the aligments
CoreGenome <- CoreAligner(Homologs[CoreSet],
                          PATH = DBPath,
                          GeneCalls = GeneCalls[1:4],
                          Verbose = TRUE)

# Create a distance matrix, that can be used to create a dendrogram
CoreDist <- DistanceMatrix(myXStringSet = CoreGenome,
                           verbose = FALSE,
                           correction = "Jukes-Cantor")

CoreDend <- IdClusters(myDistMatrix = CoreDist,
                       myXStringSet = CoreGenome,
                       method = "NJ",
                       verbose = FALSE,
                       showPlot = TRUE,
                       type = "dendrogram")
```
The presence and absence of genes across the set of genomes can also be used to create a distance matrix
```{r, echo = FALSE}
# Create a logical presence/absence matrix for orthologs
LogicalMatrix <- vector("list",
                        length = length(Homologs))
pBar <- txtProgressBar(style = 1L)
TimeStart <- Sys.time()
for (i in seq_along(Homologs)) {
  LogicalMatrix[[i]] <- vector("logical",
                               length = ncol(Homologs[[i]]))
  for (j in seq_along(LogicalMatrix[[i]])) {
    if (length(unique(Homologs[[i]][, j])) > 1L) {
      LogicalMatrix[[i]][j] <- TRUE
    }
  }
  setTxtProgressBar(pb = pBar,
                    value = i/length(LogicalMatrix))
}

LogicalMatrix <- do.call(cbind, LogicalMatrix)

# Use this logical matrix to create a distance matrix
PanGenome <- dist(LogicalMatrix,
                  method = "binary")

# Create a dendrogram
PanDend <- IdClusters(myDistMatrix = PanGenome,
                      method = "NJ",
                      type = "dendrogram",
                      showPlot = TRUE,
                      verbose = FALSE)
```
Compare two genomes with a tangleogram using phytools
```{r}
tf1 <- tempfile()
tf2 <- tempfile()

WriteDendrogram(x =PanDend,
                file = tf1)

WriteDendrogram(x = CoreDend,
                file = tf2)

PanPhylo <- read.tree(file = tf1)
CorePhylo <- read.tree(file = tf2)

unlink(tf1)
unlink(tf2)

CompObject <- cophylo(tr1 = PanPhylo,
                      tr2 = CorePhylo)

plot(CompObject,
     link.type = "curved",
     link.lwd = 3,
     link.lty = "solid",
     link.col = make.transparent("blue", 0.25),
     f.size = 0.35)
```

TODO: A Lot








